A simulation model of DNA methylation data

### 2015-05-03 Rmarkdown/methsim.Rmd
Add methsim figures

Recall that $B_{i, j}$ is the underlying _true_ methylation level at locus $i$ in sample $j$. In the context of simulating data for comparing differential methylation calling methods, we want these to be group-specific. That is, we want to specify $B_{i, j^{k}}$, where $j^{k}$ indicates that sample $j$ comes from group $k$. For non-differentially methylated loci all $k$ groups have identical $B_{i, j^{k}} = B_{i, j^{0}}$; differentially methylated loci are simulated by setting $B_{i, j^{k}} \neq B_{i, j^{k'}}$ for some $k \neq k'$.

Under the beta-binomial model, the true methylation level, $B^{i, j}$, is assumed to follow a $Beta(\mu_{i, j^{k}}, \phi_{i, j^{k}})$ distribution, where $\mu_{i, j^{k}}$ is the mean and $\phi_{i, j^{k}}$ is the dispersion of the beta distribution[^beta_parameterisation]. Both the means, $\mu_{i, j^{k}}$, and the dispersions, $\phi_{i, j^{k}}$, are group-specific and are allowed to vary across methylation loci ($i$). The dispersion parameter models the within-group variability of the $B_{i, j}$, i.e. the within-group _biological variability_ of DNA methylation.

@@

An alternative to specifying a parametric distribution for the $B_{i, j}$ is to sample these from real data, e.g., by sampling some observed $\beta_{i, j}$ for a particular dataset and treating them as if they were observations on $B_{i, j}$. `WGBSSuite`\citep{Rackham:2015bv} uses a modified form of this approach. In `WGBSSuite`, a hidden Markov model is used to classify every CpG as having an underlying state ("de-methylated", "1st transition", "2nd transition" or "methylated"). Each of these four states has a region-specific average methylation level that is based on the distribution of $\beta$-values for a chosen dataset. For example, the average methylation level for all "de-methylated" regions is defined as $B_{'de-methylated'} = median(\{ \beta_{i, j}: \beta_{i, j} \in [0, 0.5) \}])$. Then, each $B_{i, j}$ is a perturbed version of this region-level average methylation, $B_{region-type}$, obtained by adding on a zero-mean Gaussian random variable, i.e. $B_{i, j} = B_{region-type} + \epsilon_{i, j}$, where $\epsilon_{i, j} \eqd Normal(0, s^{2})$. Care needs to be taken that $0 \leq B_{i, j} \leq 1$.

@@

As we have seen, the simulation of a differentially methylated locus is straightforward; for a given locus ($i$) simply vary $B_{i, j_{k}}$ across the $k$ groups. The simulation of a differentially methylated region is more complex.

@@

@@

The second step uses a method published by \citet{Zaykin:2002ko}. The transformed values, $B_{i, j_{k}}^{*}$, are created by the transformation $\mathbf{B_{j_{k}}^{*}} = 1 - \Phi{C \Phi - 1(1 - B_{j_{k}})})$, where $\mathbf{B_{j_{k}}}$ is the vector of $B_{i, j_{k}}$, $C$ is a factor of the correlation matrix $\Sigma = C C^{'}$, where $\Sigma$ is estimated from the fitted Gaussian variogram, and where $\Phi(\cdot)$ denotes the standard normal distribution function.

@@

A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e. without simulating sequencing depth \citet[e.g.,][]{Jaffe:2012gx, Chen:2013eh, Chen:2014jb, }. These models are designed for simulating microarray data and not sequencing data, since they do not include the variability due to variation in sequencing depth. Of these methods, only \citet{Jaffe:2012gx} simulate correlation amongst the $B_{i, j_{k}}$.

@@

\citet{Sofer:2013bk} sample from a dataset of 539 Illumina 450k microarrays and select a small number of "target" CpGs, which are CpGs whose methylation level is highly variable across the 539 samples. If a block does not contain a "target" then it is sampled uniformly at random from the 539 samples. However, if a block does contain a "target" then the sampling is weighted so that the "cases" are preferentially sampled from blocks with a high level of methylation at the "target" CpG and the "controls" are preferentially sampled from blocks with a low level of methylation at the "target" CpG[^cases_and_controls]. This is essentially weighted sampling of the real data to induce differential methylation.

[^cases_and_controls]: The use of "cases" and "controls" is arbitrary, as is the choice of highly methylated for "cases" and lowly methylated for "controls" at "target" CpGs.

@@

As we have seen, DNA methylation is highly heterogeneous along the genome. Nonetheless, there are clear regions of "similarity", such as the unmethylated CpG islands and long partially methylated domains. In these locally-similar regions we might hope to model DNA methylation by a simple parametric model. `methsim`, like other simulation methods, is based on the idea of segmenting the genome into 'regions of similarity', fitting a simple model to each region and then 'stitching' the results together to form the true methylome.

@@

@@

Rather than directly modelling the transition probabilities, $\mathbf{p}$, `methsim` is parameterised by a vector of marginal probabilities, $\mathbf{B} = \{ Pr(Z_{i} = 1) \}$, and a vector of odds ratios, $\mathbf{\psi} = \{ \frac{Pr_{Z_{i + 1} = 1 | Z_{i} = 1} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 0}}{Pr_{Z_{i + 1} = 1 | Z_{i} = 0} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 1}} \} = \{ \frac{Pr_{Z_{i} = 1, Z_{i + 1} = 1} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 0}}{Pr_{Z_{i} = 1 , Z_{i + 1} = 0} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 1}} \}$. We can compute the transition probabilities, $\mathbf{p}$, from $\mathbf{B}$ and $\mathbf{\psi}$.

For each 2-tuple, $(i, i + 1)$, `methsim` first constructs the joint probability matrix, $P_{i, i + 1}$ from the marginal probabilities, $B_{i}, B_{i + 1}$, and the odds ratio, $\psi_{i}$. The general form of $P_{i, i + 1}$ is shown below:

@@

Let $B_{i} = 0.7$, $B_{i + 1} = 0.6$ and $\psi_{i, i + 1} = 2$. To begin the iterative proportional fitting algorithm, form the matrix $P_{i, i + 1}^{(0)} = \Bigl(\begin{smallmatrix}
\psi & 1 \\
   1 & 1
\end{smallmatrix} \Bigr)$. This matrix has the desired cross-ratio, $\psi$, but not the desired row and column margins. At each iteration of the algorithm, iterative proportional fitting adjusts the rows and columns such that the cross-ratio remains $\psi$ while the row and column margins converge towards the desired values. The algorithm continues in this manner, forming a series of $2 \times 2$ tables that converge pointwise (and uniquely) to a $2 \times 2$ table $P_{i, i + 1}^{(*)} \equiv P_{i, i + 1}$. For our example, $P_{i, i + 1}^{(*)} = \Bigl(\begin{smallmatrix}
0.155 & 0.145 \\
0.245 & 0.455
\end{smallmatrix} \Bigr)$, to three decimal places of precision. It is easily verified that $P^{(*)}$ has the desired cross-ratio and marginal sums.

@@

To simulate values of $\mathbf{B}$ and $\mathbf{\psi}$, `methsim` uses the empirical distributions of their estimates, $\mathbf{\beta}$ and $\widehat{\mathbf{\psi}}$, in the input methylome. More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ (the empirical distribution of $\beta_{i}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$) and $\mathbf{\psi}$ is based on $\widehat{\psi}_{MH}$. We have already seen many examples of the distribution of $\widehat{\psi}_{MH}$ in Chapter \ref{chap:co-methylation}. The empirical distributions $\beta_{i | r_{i}}$ for the _ADS_ sample are shown in Figure \ref{fig:ADS_beta_by_pm_boxplots}.

@@

There are many ways to simulate values of $\mathbf{B}$ and $\mathbf{\theta} = \log{\mathbf{\psi}}$ based on these empirical distributions. I have so far only had time to explore a few simple methods and this remains ongoing work. We will discuss the results from three methods for simulating $\mathbf{B}$ (_m1_, _m2_, and _m3_) and two ways for simulating $\mathbf{\theta}$ (_A_ and _B_):

\begin{description}
  \item[\emph{m1}] All loci in each region have the same $B$, which is sampled from $\beta_{i | r_{i}}$.
  \item[\emph{m2}] All loci in each region have the same \textbf{average} $B$, which is sampled from $\beta_{i | r_{i}}$; this is then perturbed by a $Gaussian(0, \sigma_{B}^2)$ random variable. This is similar to what is done by `WGBSsuite`. Here we use $\sigma_{B}^{2} = B \times (1 - B)$. The resulting $\mathbf{B}$ is truncated so that all values lie between 0.01 and 0.99 (necessary to avoid issues in the iterative proportional fitting algorithm at the boundaries).
  %\item[\emph{m3}] $\mathbf{B}$ simulated as in \emph{m2} but with the perturbations in each region simulated from a first-order autoregressive process with coefficient equal to 0.5.
  %\item[\emph{A}] $\theta$ identically zero. This is to simulate independence of within-fragment methylation states and as a control to check that `methsim` is working as intended for simulating within-fragment co-methylation.
  %\item[\emph{B}] $\theta$ sampled from $Gaussian(\mu_{IPD}, \sigma_{IPD}^{2})$, where $\mu_{IPD}$ and $\sigma_{IPD}$ are plug-in estimates computed from the distribution of chromosome-level $\widehat{\theta_{MH}}$ (autosomes only).
\end{description}

Furthermore, we will consider simulations where the sample is 'pure' (\emph{I}) and where the sample is a mixture of sub-populations with relative frequencies __TODO: Check frequencies__ (\emph{II}). We do not explore all combinations of these factors but look at five informative combinations: _m1AI_, _m1BI, _m2BI_, _m3BI_ and _m3BII_. We compare these five models to the real _ADS_ data.

The simulation of $\mathbf{\theta} = \log{\mathbf{\psi}}$ requires further explanation. The empirical distributions of $\widehat{\mathbf{\theta}}$ will only include values for small $IPD$s since these are estimated from individual sequencing reads (see Chapter \ref{chap:co-methylation} for details). When simulating a methylome, we will encounter 2-tuples with much larger $IPD$s and the question is how to deal with these?

Again, there are several options. For example, we might assume a parametric form for the distribution of $\mathbf{\psi}$ as a function of $IPD$, such as a exponential decay towards $\theta = 0$ (corresponding to independence). Given the quality and read-lengths of the _ADS_ data, I have elected to use the empirical $\widehat{\theta}_{MH}$ for pairs with $IPD \leq 180$ and set $\mu_{IPD} = 0$ for those with $IPD > 180$. $\sigma_{IPD}$ is given by the median absolute deviation of $\sigma_{IPD}$ for $IPD \leq 180$. While obviously a gross simplification, the vast majority of $NIL = 0$ CpG pairs have an $IPD \leq 180$ and so will be unaffected by this 'independence' assumption.

There is an increase in complexity for models m1 through to m4, while m5 is there as a sanity check that the within-fragment co-methylation model is working as intended. The limitations of these models may already be clear, but we will discuss these further in Section \ref{sec:methsim_results}.

@@

`methsim` uses a simple Poisson-based method for simulating bisulfite-sequencing. The user specifies the desired read length, the average sequencing coverage[^read_length], the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome. We will use $200$ bp, single-end reads with an average coverage of $30 \times$ in what follows.

@@

The result of this process is a `SimulatedBS` object storing the ID of each read that overlaps a methylation locus, along with the corresponding genomic co-ordinates and the methylation states.

@@

Biological variation is controlled through the `MethylomeParam` object and the method with which we simulate $\mathbf{B}$ and $\mathbf{\psi}$ through the `simulate()` method. Technical variation is controlled the `WGBSParam` object. If we want two samples to be very different from one another, then we would use two different `MethylomeParam` objects at the beginning of the process. However, if we want to simulate two technical replicates, then we would use the same `WGBSParam` objects; the two samples have identical 'true' methylomes and they only diverge at the 'sequencing' step of the simulation.

We might also consider more subtle ways of introducing variation. For example, we might use the same `MethylomeParam` object but vary how we simulate $\mathbf{B}$ and $\mathbf{\psi}$ between calls to `simulate()`. This avenue remains to be explored.

@@

Much of `methsim` is written in C++, using the wonderful Rcpp package \citep{Eddelbuettel:2011to, Eddelbuettel:2013if}, which greatly speeds up the running time of key procedures. Furthermore, several steps of the simulation can be run in parallel for each chromosome. Parallelism is implemented via the `BiocParallel` Bioconductor package \citep{BiocParallel}.

The `SimulatedMethylome` and `SimulatedBS` objects are large in memory. This is to some extent unavoidable; we are simulating whole-genome bisulfite-sequencing sequencing experiments that produce an enormous amount of data.

@@

First looking at the distribution of $\beta$-values, we see that all of _m1AI_, _m1BI_, _m2BI_, _m3BI_ and _m3BII_ do a reasonable job of capturing the bimodality of $\beta$-values (Figure \ref{fig:methsim_plotBetaFreqPoly}) and their relationship to CpG islands (Figure \ref{fig:methsim_plotBetaFreqPoly_CGI}). The effect of both the independent perturbations (_m2_) and the correlated perturbations (_m3_) appear to be dominated by variation due to variation in sequencing coverage. The lack of within-fragment co-methylation in _m1AI_ does not affect the genome-wide distribution of methylation levels. The mixture model, _m3BII_, has noticeably more intermediate $\beta$-values owing to the increased heterogeneity at each CpG across the sub-populations.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly_CGI.pdf}
\caption[\texttt{methsim} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \texttt{methsim} simulated data and the \emph{ADS} input sample, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10 \times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.
\label{fig:methsim_plotBetaFreqPoly_CGI}
\end{figure}

Turning our attention to within-fragment co-methylation, Figure \ref{fig:methsim_methsim_MH_by_seqnames} shows the genome-level and chromosome-level Mantel-Haenszel estimates of within-fragment co-methylation. Figure \ref{fig:methsim_MH_by_CGI} shows the genome-level estimates of within-fragment co-methylation stratified by CpG islands. Reassuringly, we see no evidence of within-fragment co-methylation in _m1AI_, which simulates independence of within-fragment methylation states. The trend of the genome-level estimates, by design, very closely matches that of the _ADS_ input sample. We see, however, that the chromosome-level estimates of _m1BI_, _m2BI_, _m3BI_ and _m3BII_ are perhaps a little too homogeneous when compared to the _ADS_ input sample. We also see in the results for _m1BI_, _m2BI_, _m3BI_ and _m3BII_ the effect of simulating from a $Gaussian(0, \sigma)$ for $IPD > 180$, as outlined in Section \ref{subsec:parameterising_the_model}. The _m3BII_ model again stands out, this time with a noticeably higher level of within-fragment co-methylation. Each sub-population has its own $\mathbf{\theta}$ and, although these are sampled from the same parametric model, this suggests that such heterogeneity in the sample will artificially inflate estimates of within-fragment co-methylation, even those made at the genome-level. While this requires further investigation, it raises the possibilit that increased genome-level estimates of within-fragment co-methylation may in fact be measuring increased heterogeneity of the sample.

@@

Figures \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf} shows the overwhelming nature of this modelling decision. When compared to the correlations of $\beta$-values from the _ADS_ input sample, we see that in all five models the correlations of $\beta$-values are far too strong. Stratifying these correlations by CpG islands, we see that the source of this problem is in regions outside of CpG islands (Figure \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}). This gives some hope that this deficiency may fixed by a more careful modelling of $\beta$-values outside CpG islands, and remains a source of ongoing work.

>>>>

Recall that $B_{i, j}$ is the underlying _true_ methylation level at locus $i$ in sample $j$. In the context of simulating data for comparing differential methylation calling methods, we want these to be group-specific. That is, we want to specify $B_{i, j_{k}}$, where $j_{k}$ indicates that sample $j$ comes from group $k$. For non-differentially methylated loci all $k$ groups have identical $B_{i, j_{k}} = B_{i, j^{0}}$; differentially methylated loci are simulated by setting $B_{i, j_{k}} \neq B_{i, j^{k'}}$ for some $k \neq k'$.

Under the beta-binomial model, the true methylation level, $B_{i, j}$, is assumed to follow a $Beta(\mu_{i, j_{k}}, \phi_{i, j_{k}})$ distribution, where $\mu_{i, j_{k}}$ is the mean and $\phi_{i, j_{k}}$ is the dispersion of the beta distribution[^beta_parameterisation]. Both the means, $\mu_{i, j_{k}}$, and the dispersions, $\phi_{i, j_{k}}$, are group-specific and are allowed to vary across methylation loci ($i$). The dispersion parameter models the within-group variability of the $B_{i, j}$, i.e. the within-group _biological variability_ of DNA methylation.

@@

An alternative to specifying a parametric distribution for the $B_{i, j}$ is to sample these from real data, e.g., by sampling some observed $\beta_{i, j}$ for a particular dataset and treating them as if they were observations on $B_{i, j}$. `WGBSSuite`\citep{Rackham:2015bv} uses a modified form of this approach. In `WGBSSuite`, a hidden Markov model is used to classify every CpG as having an underlying state ("de-methylated", "1st transition", "2nd transition" or "methylated"). Each of these four states has a region-specific average methylation level that is based on the distribution of $\beta$-values for a chosen dataset. For example, the average methylation level for all "de-methylated" regions is defined as $B_{de-methylated} = median(\{ \beta_{i, j}: \beta_{i, j} \in [0, 0.5) \}])$. Then, each $B_{i, j}$ is a perturbed version of this region-level average methylation, $B_{region-type}$, obtained by adding on a zero-mean Gaussian random variable, i.e. $B_{i, j} = B_{region-type} + \epsilon_{i, j}$, where $\epsilon_{i, j} \eqd Normal(0, s^{2})$. Care needs to be taken that $0 \leq B_{i, j} \leq 1$.

@@

As we have seen, the simulation of a differentially methylated locus is straightforward; for the $i^{th}$ locus, simply vary $B_{i, j_{k}}$ across the $k$ groups. The simulation of a differentially methylated region is more complex.

@@

\FloatBarrier

@@

The second step uses a method published by \citet{Zaykin:2002ko}. The transformed values, $B_{i, j_{k}}^{*}$, are created by the transformation $\mathbf{B_{j_{k}}^{*}} = 1 - \Phi \Big \{ C \Phi^{-1}(1 - B_{j_{k}}) \Big \}$, where $\mathbf{B_{j_{k}}}$ is the vector of $B_{i, j_{k}}$, $C$ is a factor of the correlation matrix $\Sigma = C C^{'}$, where $\Sigma$ is estimated from the fitted Gaussian variogram, and where $\Phi(\cdot)$ denotes the standard normal distribution function.

@@

A separate class of model-based simulation methods are those that simulate the $\beta$-values directly, i.e. without simulating sequencing depth \citet[e.g.,][]{Jaffe:2012gx, Chen:2013eh, Chen:2014jb}. These models are designed for simulating microarray data and not sequencing data, since they do not include the variability due to variation in sequencing depth. Of these methods, only \citet{Jaffe:2012gx} simulate correlation amongst the $B_{i, j_{k}}$.

@@

\citet{Sofer:2013bk} sample from a dataset of 539 Illumina 450k microarrays and select a small number of "target" CpGs, which are CpGs whose methylation level is highly variable across the 539 samples. If a block does not contain a target then it is sampled uniformly at random from the 539 samples. However, if a block does contain a target then the sampling is weighted so that the "cases" are preferentially sampled from blocks with a high level of methylation at the target CpG and the "controls" are preferentially sampled from blocks with a low level of methylation at the target CpG[^cases_and_controls]. This is essentially weighted sampling of the real data to induce differential methylation.

[^cases_and_controls]: The use of "cases" and "controls" is arbitrary, as is the choice of highly methylated for "cases" and lowly methylated for "controls" at target CpGs.

@@

As we have seen, DNA methylation is highly heterogeneous along the genome. Nonetheless, there are clear regions of `similarity', such as the unmethylated CpG islands and long partially methylated domains. In these locally-similar regions we might hope to model DNA methylation by a simple parametric model. `methsim`, like other simulation methods, is based on the idea of segmenting the genome into 'regions of similarity', fitting a simple model to each region and then 'stitching' the results together to form the true methylome.

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_by_pm_boxplots.pdf}
\caption[\emph{EPISCOPE}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{EPISCOPE} dataset.}
\label{fig:EPISCOPE_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Lister_beta_by_pm_boxplots.pdf}
\caption[\emph{Lister}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Lister} dataset.}
\label{fig:Lister_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_beta_by_pm_boxplots.pdf}
\caption[\emph{Seisenberger}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Seisenberger} dataset.}
\label{fig:Seisenberger_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_beta_by_pm_boxplots.pdf}
\caption[\emph{Ziller}: $\beta$-values my region-type]{Boxplots of CpG $\beta$-values in each region-type following segmentation by \texttt{MethylSeekR} for the \emph{Ziller} dataset.}
\label{fig:Ziller_beta_by_pm_boxplots}
\end{figure}

@@

Rather than directly modelling the transition probabilities, $\mathbf{p}$, `methsim` is parameterised by a vector of marginal probabilities, $\mathbf{B} = \{ Pr(Z_{i} = 1) \}$, and a vector of odds ratios, $\bm{\psi} = \frac{Pr(Z_{i + 1} = 1 | Z_{i} = 1) \times Pr(Z_{i + 1} = 0 | Z_{i} = 0)}{Pr(Z_{i + 1} = 1 | Z_{i} = 0) \times Pr(Z_{i + 1} = 0 | Z_{i} = 1)} = \frac{Pr(Z_{i} = 1, Z_{i + 1} = 1) \times Pr(Z_{i} = 0 , Z_{i + 1} = 0)}{Pr(Z_{i} = 1 , Z_{i + 1} = 0) \times Pr(Z_{i} = 0 , Z_{i + 1} = 1)}$. We can compute the transition probabilities, $\mathbf{p}$, from $\mathbf{B}$ and $\bm{\psi}$.

For each 2-tuple, $(i, i + 1)$, `methsim` first constructs the joint probability matrix, $P_{i, i + 1}$ from the marginal probabilities, $B_{i}$ and $B_{i + 1}$, and the odds ratio, $\psi_{i}$. The general form of $P_{i, i + 1}$ is shown below:

@@

Let $B_{i} = 0.7$, $B_{i + 1} = 0.6$ and $\psi_{i} = 2$. To begin the iterative proportional fitting algorithm, form the matrix $P_{i, i + 1}^{(0)} = \Bigl(\begin{smallmatrix}
\psi_{i} & 1 \\
   1 & 1
\end{smallmatrix} \Bigr)$. This matrix has the desired cross-ratio, $\psi_{i}$, but not the desired row and column margins. At each iteration of the algorithm, iterative proportional fitting adjusts the rows and columns such that the cross-ratio remains $\psi_{i}$ while the row and column margins converge towards the desired values. The algorithm continues in this manner, forming a series of $2 \times 2$ tables that converge pointwise (and uniquely) to a $2 \times 2$ table $P_{i, i + 1}^{(*)} \equiv P_{i, i + 1}$. For our example, $P_{i, i + 1}^{(*)} = \Bigl(\begin{smallmatrix}
0.155 & 0.145 \\
0.245 & 0.455
\end{smallmatrix} \Bigr)$, to three decimal places of precision. It is easily verified that $P^{(*)}$ has the desired cross-ratio and marginal sums.

@@

To simulate values of $\mathbf{B}$ and $\bm{\psi}$, `methsim` uses the empirical distributions of their estimates, $\bm{\beta}$ and $\widehat{\bm{\theta}} = \log(\widehat{\bm{\psi}})$, in the input methylome. More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ (the empirical distribution of $\beta_{i}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$) and $\bm{\theta}$ is based on $\widehat{\theta}_{MH}$. We have already seen many examples of the distribution of $\widehat{\theta}_{MH}$ in Chapter \ref{chap:co-methylation}. The empirical distributions $\beta_{i | r_{i}}$ for the _ADS_ sample are shown in Figure \ref{fig:ADS_beta_by_pm_boxplots}.

@@

There are many ways to simulate values of $\mathbf{B}$ and $\bm{\theta} = \log{\bm{\psi}}$ based on these empirical distributions. I have so far only had time to explore a few simple methods and this remains ongoing work. We will discuss the results from three methods for simulating $\mathbf{B}$ (_m1_, _m2_, and _m3_) and two ways for simulating $\bm{\theta}$ (_A_ and _B_):

\begin{description}
  \item[\emph{m1}] All loci in each region have the same $B$, which is sampled from $\beta_{i | r_{i}}$.
  \item[\emph{m2}] All loci in each region have the same \textbf{average} $B$, which is sampled from $\beta_{i | r_{i}}$; this average is then perturbed for each locus by a $Gaussian(0, \sigma_{B}^2)$ random variable. This is similar to what is done by \texttt{WGBSsuite}. Here I have used $\sigma_{B}^{2} = B \times (1 - B)$. The resulting $\mathbf{B}$ is truncated so that all values lie between 0.01 and 0.99 (necessary to avoid issues in the iterative proportional fitting algorithm at the boundaries).
  %\item[\emph{m3}] $\mathbf{B}$ simulated as in \emph{m2} but with the perturbations in each region simulated from a first-order autoregressive process with coefficient equal to 0.5.
  %\item[\emph{A}] $\theta$ identically zero. This is to simulate independence of within-fragment methylation states and as a control to check that `methsim` is working as intended for simulating within-fragment co-methylation.
  %\item[\emph{B}] $\theta$ sampled from $Gaussian(\mu_{IPD}, \sigma_{IPD}^{2})$, where $\mu_{IPD}$ and $\sigma_{IPD}$ are plug-in estimates computed from the distribution of chromosome-level $\widehat{\theta_{MH}}$ (autosomes only).
\end{description}

Furthermore, we will consider simulations where:

\begin{description}
  \item[\emph{I}] The sample is 'pure' and all haplotypes have the same parameters.
  \item[\emph{II}] The sample is a mixture of subpopulations with different parameters. We will consider a sample with four subpopulations with relative frequencies equal to $(0.6, 0.25, 0.1, 0.05)$.
\end{description}

We do not explore all combinations of these factors in the results presented here but look at five informative combinations: _m1AI_, _m1BI_, _m2BI_, _m3BI_ and _m3BII_. We compare these five models to the real _ADS_ data. The limitations of these models may already be clear, and we will discuss these further in Section \ref{sec:methsim_results}.

The simulation of $\bm{\theta} = \log{\bm{\psi}}$ requires further explanation. The empirical distributions of $\widehat{\bm{\theta}}$ will only include values for small $IPD$s since these are estimated from individual sequencing reads (see Chapter \ref{chap:co-methylation} for details). When simulating a methylome, we will encounter 2-tuples with much larger $IPD$s and the question is how to deal with these.

Again, there are several options. For example, we might assume a parametric form for the distribution of $\bm{\theta}$ as a function of $IPD$, such as a exponential decay towards $\theta = 0$ (corresponding to independence). Given the quality and read-lengths of the _ADS_ data, I have elected to use the empirical $\widehat{\theta}_{MH}$ for pairs with $IPD \leq 180$ and set $\mu_{IPD} = 0$ for those with $IPD > 180$. $\sigma_{IPD}$ is given by the median absolute deviation of $\sigma_{IPD}$ for pairs with $IPD \leq 180$. While obviously a gross simplification, the vast majority of CpG pairs with $NIL = 0$ have an $IPD \leq 180$ and so will be unaffected by this 'independence' assumption.

\FloatBarrier

@@

`methsim` uses a simple Poisson-based method for simulating bisulfite-sequencing. The user specifies the desired read length, the average sequencing coverage[^read_length], the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome. We will use $200$ bp, single-end reads (equivalently, $100$ bp paired-end reads that are always end-to-end) with an average coverage of $30 \times$ in what follows.

@@

The result of this process is a `SimulatedWGBS` object storing the ID of each read that overlaps a methylation locus, along with the corresponding genomic co-ordinates and the methylation states.

@@

Biological variation is controlled through the `MethylomeParam` object and the method with which we simulate $\mathbf{B}$ and $\bm{\psi}$ through the `simulate()` method. Technical variation is controlled the `WGBSParam` object. If we want two samples to be very different from one another, then we would use two different `MethylomeParam` objects at the beginning of the process. However, if we want to simulate two technical replicates, then we would use the same `WGBSParam` objects; the two samples have identical 'true' methylomes and they only diverge at the 'sequencing' step of the simulation.

We might also consider more subtle ways of introducing variation. For example, we might use the same `MethylomeParam` object but vary how we simulate $\mathbf{B}$ and $\bm{\psi}$ between calls to `simulate()`. This avenue remains to be explored.

@@

While `methsim` is an R package, much of it is written in C++, using the Rcpp package \citep{Eddelbuettel:2011to, Eddelbuettel:2013if}, which greatly speeds up the running time of key procedures. Furthermore, several steps of the simulation can be run in parallel for each chromosome. Parallelism is implemented via the `BiocParallel` Bioconductor package \citep{BiocParallel}.

The `SimulatedMethylome` and `SimulatedWGBS` objects are large in memory. This is to some extent unavoidable; we are simulating whole-genome bisulfite-sequencing sequencing experiments that produce an enormous amount of data.

@@

Firstly, looking at the distribution of $\beta$-values, we see that all of _m1AI_, _m1BI_, _m2BI_, _m3BI_ and _m3BII_ do a reasonable job of capturing the bimodality of $\beta$-values (Figure \ref{fig:methsim_plotBetaFreqPoly}) and their relationship to CpG islands (Figure \ref{fig:methsim_plotBetaFreqPoly_CGI}). The effect of both the independent perturbations (_m2_) and the correlated perturbations (_m3_) appear to be dominated by variation due to variation in sequencing coverage. The lack of within-fragment co-methylation in _m1AI_ does not affect the genome-wide distribution of methylation levels. The mixture model, _m3BII_, has noticeably more intermediate $\beta$-values owing to the increased heterogeneity at each CpG across the sub-populations.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_plotBetaFreqPoly_CGI.pdf}
\caption[\texttt{methsim} $\beta$-value kernel density stratified by CpG islands]{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the \texttt{methsim} simulated data and the \emph{ADS} input sample, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10 \times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:methsim_plotBetaFreqPoly_CGI}
\end{figure}

Turning our attention to within-fragment co-methylation, Figure \ref{fig:methsim_MH_by_seqnames} shows the genome-level and chromosome-level Mantel-Haenszel estimates of within-fragment co-methylation. Figure \ref{fig:methsim_MH_by_CGI} shows the genome-level estimates of within-fragment co-methylation stratified by CpG islands. These results are only available for CpG pairs with $NIL = 0$ because the `asMethPat()` function does not create $NIL > 0$ pairs.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_MH_by_seqnames.pdf}
\caption[\texttt{methsim}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \texttt{methsim} simulated data and the \emph{ADS} input sample using $NIL = 0$ pairs. The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA.}
\label{fig:methsim_MH_by_seqnames}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_MH_by_CGI.pdf}
\caption[\texttt{methsim}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \texttt{methsim} simulated data and the \emph{ADS} input sample using $NIL = 0$ pairs. Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI).}
\label{fig:methsim_MH_by_CGI}
\end{figure}

Reassuringly, we see no evidence of within-fragment co-methylation in _m1AI_, which simulates independence of within-fragment methylation states. The trend of the genome-level estimates, by design, very closely matches that of the _ADS_ input sample. We see, however, that the chromosome-level estimates of _m1BI_, _m2BI_, _m3BI_ and _m3BII_ are perhaps a little too homogeneous when compared to the _ADS_ input sample. We also see in the results for _m1BI_, _m2BI_, _m3BI_ and _m3BII_ the effect of simulating from a $Gaussian(0, \sigma)$ for $IPD > 180$, as outlined in Section \ref{subsec:parameterising_the_model}. The _m3BII_ model again stands out, this time with a noticeably higher level of within-fragment co-methylation. Each sub-population has its own $\bm{\theta}$ and, although these are sampled from the same parametric model, this suggests that such heterogeneity in the sample will artificially inflate estimates of within-fragment co-methylation, even those made at the genome-level. While this requires further investigation, it raises the possibilit that increased genome-level estimates of within-fragment co-methylation may in fact be measuring increased heterogeneity of the sample.

@@

Figures \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI} shows the overwhelming nature of this modelling decision. When compared to the correlations of $\beta$-values from the _ADS_ input sample, we see that in all five models the correlations of $\beta$-values are far too strong. Stratifying these correlations by CpG islands, we see that the source of this problem is in regions outside of CpG islands (Figure \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}). This gives some hope that this deficiency may fixed by a more careful modelling of $\beta$-values outside CpG islands, and remains a source of ongoing work.
